---
title: "Basics Tutorial"
author: "Jordan Wylie"
date: "10/7/2020"
output:
  html_document
---

# Basics of R Tutorial (R Markdown Executable File - knits to HTML)

This tutorial is to help new users get started with R with the goal of analyzing data. This tutorial was specifically made for undergraduate research assistants, but should be helpful for anyone looking to get started with data analysis in R. 

If your goal is to become literate in R programming language, there are many other resources better suited to that goal. 

For example:

- https://swirlstats.com/students.html

- https://www.datacamp.com/courses/introduction-to-the-tidyverse

- https://kiirstio.wixsite.com/kowen/post/the-25-days-of-christmas-an-r-advent-calendar

This was created in RStudio (using RMarkdown). 

R studio is an Integrated Development Environment (IDE) for R, you will need to download both R and RStudio

Here is the link to R: https://cran.rstudio.com/

Here is the link to R Studio: https://rstudio.com/products/rstudio/download/#download

# Getting started

The last portion of this tutorial (Section 6) utilizes a dataset that is available on OSF. You can access it by request here: https://osf.io/3xmny/. This is a work in progress. If you have any questions, feedback, or concerns, please email me! 

Before we dive into the data. We will cover a few basics of the language. Again, this is not exhaustive as covering all the parts of the programming language is outside of the purview of this tutorial, and resources to do that are readily available all over. 

One short note: When you use the **#** in a code block (or in your .r file), this allows you to add comments to your code. This makes life a whole lot easier because it is always true that your future self has no idea what your past self was thinking or doing. Seriously. Use comments, your future self will thank you. Additionally, R is case sensitive. Most of the time, if your code won't work, there is a typo of case error hidden somewhere!

**0. The Set-up **

This section has some set up functions that will be useful. For example, using: `data <- read.csv(FILEHERE)` and then `colnames(data)` would take a .csv and read it into R, then pull the variable names for you. I find this part to be less intuitive, so if you are like me, you will get used to it with practice.

```{r setup, include=TRUE, echo=TRUE, results='hide', warning=FALSE, message=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

#Packages: R is so useful for data analysis because many of the things you need to run data analyses are already created for you in packaged. As such, you will rely heavily on them since a bunch of great developers have already put the time in to create these tools. 

# IMPORTANT: All of this is open source and on CRAN, so when you open the install a package, it will search for it there (there are other ways of gettign packages, but that is not something you will liekly need to do)

# You have to install the packages once to move them from CRAN to your R - so you can use them!

#install.packages(“readr”) # Will install the default location
#install.packages(“car”)
#install.packages(“tidyverse”)
#install.packages(“dplyr”)
#install.packages(“psych”)
#install.packages(“ggplot2”)
#install.packages(“plyr”)

#After you have downloaded the package, you have to use library to read it in. So- you download it from CRAN one time. But EVERYTIME you need to use the package from your computer, you need to use library. Also, sometimes you will see that the package contains a function that is the same name as a function from a different package. In this case, you will see a message that alerts you that the most recently uploaded package is going to mask the previous one! There are ways to get around this. You can call the package in your code to make sure it pulls the right version (or you can unload the package). There will be an example of this in Section 7. 

library(readr)
library(utils)
library(car)
library(tidyverse)
library(dplyr)
library(psych)
library(ggplot2)
library(plyr)
library(summarytools)
library(lmtest)

#To set up your the working directory, you can use setwd("C:/Users/NAME/LOCATION/"). But often times you will need to use `Sys.getenv("HOME")` to find where Home base is. 

# You can remove the `HOME` directory from paths, and replace it with `~`
#setwd("~/Documents")

# Loading data - You can access files from your working directory.

# CSV (best way to save files!)
#myData <- read.csv("FILENAME.csv")

# SPSS files! 
library(foreign)
#MySPSSdata<- read.spss("FILENAME.sav")

## You can also load data straight from OSF, but we won't cover that right now

#Downloading what we will need for the exercise at the end. 
setwd("/Users/jordanwylie/Desktop/")   # Import data (csv file)/set this to wherever you downloaded the data from OSF.

dfAesth <- read.csv("Moral_Art_CleanRecode_VOSF.csv", header = T)
colnames(dfAesth)

#Saving data - when you do things in R, they are not being written directly into your datafile. This is actually really great! You will probably make (many) mistakes. Having your file untouched is useful so that you don't permanently mess anything up. But, if you want to save new variables you created or something like that, you can! I recommend that you do not overwrite the original file, but create a new one.

# CSV
#write.csv(myData, "FILENAME.csv")

```


**1. The Calculator**

You can use R like it's just a big calculator

```{r, include=TRUE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# Addition
2 + 2
# Subtraction
2 - 2
# Multiplication
2 * 3
# Division
2 / 3
# Exponents
2 ^ 3
# More complex operations
(2 + 2) / 3
2 + (2 / 3)

#Other opeations

# Square root
sqrt(9) 
sqrt(x = 9) #These are equivalent, in the first, the name is implicit

# Round a number
round(x = 1.6180339887, digits = 0)     # Including the two arguments
round(1.6180339887, 0)                  # Equivalent, names implicit
round(1.6180339887)                     # With a default (where second argument is assumed to be 0-you should always check on this to make sure you know what arguments are being passed by defaults, and which ones you need to specify)

round(3.6760698786, 4)                  # Round Phi to the 4th decimal place

# log
log(2)              # Natural log
log(2, base = 10)   # Log base 10

# e^x
exp(2)
exp(0.6931472)

# Absolute value
abs(-2)
abs(2)

#Nested oepration
exp(log(2))     # These generally operate by requiring reading from the inside-out.

```

**2. The Logician **

You can use R to evaluate logical statements

```{r, include=TRUE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

#   ==  equality
#   !=  inequality
#   >   greater than
#   >=  greater than or equal to
#   <   less than
#   <=  less than or equal to
#
#   &   AND
#   |   OR
#   !   NOT

# Equality
2 + 2 == 5      # FALSE

# Inequality
2 + 2 != 5      # TRUE
5 != 1          # TRUE

# Greater than
5 > 1           # TRUE

# Greater than or equal to
5 >= 10         # FALSE

# Less than
5 < 1           # FALSE

# Less than or equal to
4 <= 5          # TRUE

```

**3. Object assignment **

Almost every operation in R uses objects. Objects are used to store information, and often times, you will be able to think about objects as variables. But objects are really flexible (functions, for example, can be objects). 

The pieces of information in objects are called "elements". You "assign" data to objects using the assignment operator **<-**. But, you can also asign using the **=** sign (it is just not preferred). If you want others to be able to read your code (i.e., reproducible code), you should use **<-** to keep things clear. 

```{r, include=TRUE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

myNumber <- 10
print(myNumber)
myNumber #does the same thing as print

myNumber /  2   
myNumber == 10  

myNumber <- 11
myNumber == 10  #my number was overwritten when I passed the new number to the object

#You can use any letter or word to represent your variable, list, object, etc. However, you want ot stay away from letters that have meaning already in math (this is true in MATLAB as well - i, j); for example, e already has a meaning, so you shouldn't use that. 

### Vectors - vectors have magnitude and location
myObject <- c(1, 2, 9, 10)
myObject
myObject /  2 

# You can also give the elements in the object myVector names
myNewObject <- c("time 1" = 1, "time 2" = 2, "time 3" = 9, "time 4" = 10)
myNewObject

# You can select and extract specific elements from objects in many ways, most basic is using square brackets []
myObject[2]
myNewObject[2]

# You can use : and c(X,X) to pull out a subset as well
myObject[1:3]
myNewObject[c(1,2)]
myObject[c(1,4)]
myNewObject[c(1,4)]

### Vectors - vectors have magnitude and location. You can concatinate objects into a vector by forcing them together

myVector <- c(myObject, myNewObject)
myVector
```


**3. More complex objects **

You can use vectors (#s with magnitude and direction), factors (categorical variable, can have diff levels), matrices (vectors with dimension), lists (object with many types (numbers, strings,vectors,etc)), and data frames (a two dimensional data structure in R), and more!!. We will explore a few of those below. 

Note: Sometimes, you will need to figure out what object you created (so that you can change it or figure out if it will work in a particular function). To figure out the "class" of the object, you can use `class()` and `str()`. 

```{r, include=TRUE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

## Matrices + plus some manipualtion (to bind the rows = rbind(); to bind the columns = cbind())

mat1 <- matrix(1, nr = 2, nc = 2)
mat2 <- matrix(2, nr = 2, nc = 2)
rbind(mat1, mat2)
cbind(mat1, mat2)

## Lists - Lists are really useful because their elements are other objects. So, each element is its own object. Lists can get complicated and even contain other lists. We will only briefly cover lists here.

# Combine as elements in a list
myList <- list(myObject, myNewObject)
class(myList)
myList[[2]]

# The last class we will cover are dataframes. But we will come back to that in a bit!
```

**4. Generating variables & functions **

We will not go over much more of the basics, here is a brief intro into variable creation and functions below: 

```{r, include=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

# Create variables: 
varname <- 1:30
varname1 <- seq(1, 5, 0.5) 
# first number indicates the beginning of the sequence, the second one the end, and the third one the increment to be used to generate the sequence
varname2 <- seq(length=9, from=1, to=5)
varname3 <- rep(1, 30) #repeates a # for a length
varname4 <- rnorm(45, mean=1, sd=1) #Gaussian (normal) random numbers

#Create a function: 

#function (argist) {body}
species <- c("swallow", "bluebird", "cardinal") 
#This is a list of string variables (much like the names of levels of a categorical variable!))
myfun <- function(thing)
{
  print(thing)
}
myfun(species)

```


**5. Dataframes **

Dataframes arrange columns together so that they look like a spreadsheet would look. They are incredibly useful for data analysis for this reason and are what you will liekly be using the most. Before we move on to new data, we will explore dataframes using mtcars -- one of R's built-in datasets.

```{r, include=TRUE, results='hide', warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

class(mtcars)
head(mtcars)     # Displays 1st 6 rows 
tail(mtcars)     # Displays last 6 rows
colnames(mtcars) # Get the variable names
# View(mtcars)     # Opens the dataset so that you can preview it in RStudio 

#Explore 
mtcars[1]       # Single bracket keeps the rownames
mtcars[[1]]     # Double bracket just prints the elements

str(mtcars[1])
str(mtcars[[1]])

mtcars["mpg"]       # Single bracket keeps the rownames
mtcars[["mpg"]]     # Double bracket just prints the elements

str(mtcars["mpg"])
str(mtcars[["mpg"]])

## Using matrix notation
mtcars[ ,1]     # [rows, columns]
str(mtcars[ ,1])

# Matrix notation with names
mtcars[ ,"mpg"]     # [rows, columns]
str(mtcars[ ,"mpg"])

mtcars$mpg          # Just prints elements
str(mtcars$mpg)

# Matrix notation also means that you can access rows
mtcars[1, ]     # [rows, columns]
str(mtcars[1, ])

# By row names
mtcars["Mazda RX4", ]   
str(mtcars["Mazda RX4", ])

```


**6. Applying these concepts **

We will now explore some pilot data using what we've learned. One reason to do this is because often times the challenge for reserachers is not in learning basic coding, but actually applying it to real data that is messy. These data are not published yet and should not be shared without permission. 

```{r, include=TRUE, echo=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

# GETTING STARTED W DATA
# CodeBook:Gender: 1 = male, 2 = female #

######## Load Necessary Packages / Load Data ####
# Packages - you must use "install.packages("[insert name of package]") if you do not already have the package
# Use ??functionName if you need to figure out what package a function call comes from

#install.packages("readr") #example if you don't have a package installed

#We already created our dataframe and called it dfAesth. We also already installed our packages. However, below I have an example how you might grab data off your local drive again.

#setwd("/Users/jordanwylie/Desktop/")      # Import data (csv file)
#dfAesth <- read.csv("DATANAME.csv", header = T)
#colnames(dfAesth) #This is commented out to reduce space, but this would give you the variable names!

##################################

# Recode Data #
dfAesth$Gender <- ifelse(dfAesth$gender == 1, "Male",
                         ifelse(dfAesth$gender == 2, "Female", NA))
dfAesth$Gender  <- as.factor(dfAesth$Gender)

##################################################################################

#General Demographics Analyses
plyr::count(dfAesth$Gender)

##################################
#Create the first IV
#We will check the alpha
OCDr <- dfAesth[,c("OCD_1","OCD_2","OCD_3", 
                  "OCD_4", "OCD_5","OCD_6",
                  "OCD_7", "OCD_8", "OCD_9",
                  "OCD_10", "OCD_11", "OCD_12",
                  "OCD_13", "OCD_14", "OCD_15", 
                  "OCD_16", "OCD_17", "OCD_18")]  
# psych::alpha(OCDr) #These are commented out because the output is long. But this will give you internal reliability 
##################################
#Then we will create the index 
dfAesth$OCDIndex <- as.numeric(rowSums(dfAesth[,c("OCD_1","OCD_2","OCD_3", 
                                                "OCD_4", "OCD_5","OCD_6",
                                                "OCD_7", "OCD_8", "OCD_9",
                                                "OCD_10", "OCD_11", "OCD_12",
                                                "OCD_13", "OCD_14", "OCD_15", 
                                                "OCD_16", "OCD_17", "OCD_18")])) 

#Create the DVs
LikingBP <- dfAesth[,c("NewLiking_BP_1","NewLiking_BP_2","NewLiking_BP_3", 
                   "NewLiking_BP_4", "NewLiking_BP_5","NewLiking_BP_6",
                   "NewLiking_BP_7", "NewLiking_BP_8", "NewLiking_BP_9",
                   "NewLiking_BP_10")]
# psych::alpha(LikingBP) #internal reliability
##################################

dfAesth$LikingBP <- as.numeric(rowMeans(dfAesth[,c("NewLiking_BP_1","NewLiking_BP_2","NewLiking_BP_3", "NewLiking_BP_4", "NewLiking_BP_5","NewLiking_BP_6", "NewLiking_BP_7", "NewLiking_BP_8", "NewLiking_BP_9","NewLiking_BP_10")]))

LikingBN <- dfAesth[,c("NewLiking_BN_1","NewLiking_BN_2","NewLiking_BN_3", 
                       "NewLiking_BN_4", "NewLiking_BN_5","NewLiking_BN_6",
                       "NewLiking_BN_7", "NewLiking_BN_8", "NewLiking_BN_9",
                       "NewLiking_BN_10")]
# psych::alpha(LikingBN) #internal reliability
##################################

dfAesth$LikingBN <- as.numeric(rowMeans(dfAesth[,c("NewLiking_BN_1","NewLiking_BN_2","NewLiking_BN_3","NewLiking_BN_4","NewLiking_BN_5","NewLiking_BN_6", "NewLiking_BN_7","NewLiking_BN_8", "NewLiking_BN_9", "NewLiking_BN_10")]))

```

```{r, include=TRUE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
##################################
#SUMMARIES

# The operator %>% basically tells R to take the value of that which is to the left and 
# pass it to the right as an argument

# detach(plyr) 
#This is how you would remove a package that you don't want to mask something

aesth.summaryBN <- dfAesth %>% # the names of the new data frame and the data frame to be summarized
  dplyr::group_by(Gender) %>%   #the grouping variable
  dplyr::summarise(mean_summBN = mean(LikingBN),  #calculates the mean by group
            sd_summBN = sd(LikingBN), #calculates the sd by group
            SE_summBN = parameters::standard_error(LikingBN)) # calculates se by group
aesth.summaryBN <- aesth.summaryBN[-c(3),] #remove the NA row
aesth.summaryBN

aesth.summaryBP <- dfAesth %>% 
 dplyr::group_by(Gender) %>%   
 dplyr::summarise(mean_summBP = mean(LikingBP), 
            sd_summBP = sd(LikingBP), 
            SE_summBP = parameters::standard_error(LikingBP))
aesth.summaryBP <- aesth.summaryBP[-c(3),] #remove the NA row
aesth.summaryBP

##################################
#CORRELATIONS
corVar1 <- dfAesth[,c("LikingBP","LikingBN","OCDIndex")]
corrVarMat <- lowerCor(corVar1)
corr.test(corVar1) # use = "pairwise",method="pearson",adjust="holm", default is "pairwise"

##################################
#REGRESSIONS
model1 <- lm(LikingBP ~ OCDIndex, data = dfAesth)
summary(model1)
model2 <- lm(LikingBN ~ OCDIndex, data = dfAesth)
summary(model2)


```

**7. Visualizations **

Here we will go thru creation of a basic scatterplot. All code PRECEDES the respective figure.

```{r scatter1, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
#Create a scatterplot with OCD on the x-axis and Liking BP on the y-axis

# remove the one person who selected gender other than male and female for visualization
dfAesth <- dfAesth[!is.na(dfAesth$Gender), ]

#Example: 
scatter1 <- ggplot(aes(OCDIndex, LikingBP), data = dfAesth)
scatter1 + geom_point(aes(colour = Gender), shape = 16, size = 3) + labs(x = "OCD Score", y = "Liking Judgments")

#Add a regression line to the scatterplot:
scatter1 + geom_point (shape = 16, size = 3) + 
  geom_smooth(method = "lm") + 
  labs(x = "OCD Score", y = "Liking Judgments")

#Add the 95% confidence interval
scatter1 + geom_point (shape = 16, size = 3) + 
  geom_smooth(method = "lm", colour = "blue", fill = "red") + 
  labs (x = "OCD Score", y = "Liking Judgments")

```



**7.1 Visualizations **

Adding regression lines by Gender. 
```{r scatter2, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

#Create a new scatterplot with separate regression lines for women and men:
scatter2 <- ggplot(aes(OCDIndex, LikingBP), data = dfAesth) + 
  geom_point () + 
  geom_smooth(method = "lm", aes(fill = Gender), colour="black", size=0.5) + 
  labs(x = "OCD Score", y = "Liking Judgments")
scatter2

```



**7.2 Visualizations - Histograms **

Creating a histogram with the mean included. 
```{r histogram1, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

#Create a histogram showing the distribution of liking. 
histogram1 <- ggplot(aes(LikingBP), data = dfAesth) + 
  geom_histogram(binwidth = 1, color="black", fill="white") + 
  labs(x = "LikingBP", y = "Count")

#Add a line for the mean to the histogram:
histogram1 + geom_vline(aes(xintercept=mean(LikingBP)), color="blue", linetype="dashed", size=1)

```







**7.2.1 Visualizations **

Histogram grouped by Gender. 

```{r histogram2, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

#Group histogram by gender:

histogram2 <- ggplot(aes(OCDIndex), data = dfAesth) + geom_histogram(binwidth = 1, color="black", fill="white") + 
  labs(x = " OCD Score ", y = "Count")

histogram2 <- ggplot(aes(OCDIndex, colour = Gender, fill = Gender), data = dfAesth) + 
  geom_histogram(binwidth = 1, position="dodge") + 
  labs(x = "OCD Score", y = "Count")  

histogram2
```
